A Short and Mechanized Consistency Proof for Dependent Type Theory

### 2024-02-13 paper/paper.tex
Add text about the normal form extension

@@

\subsection{Toward Decidability of Type Conversion and $\eta$ laws}
In this section, we sketch out how the logical relation can be extended
show the existence of $\beta\eta$ normal forms for well-typed terms.
In other words, our goal is to show that the following property holds
in our system.

>>>>

\section{Toward Decidability of Type Conversion and $\eta$ laws}
In this section, we sketch out how the logical relation from
Section~\ref{sec:logreldep} can be extended show the existence of
$\beta\eta$ normal forms for (open and closed) well-typed terms.

The updated logical relation and its auxiliary definitions are shown
in Figure~\ref{fig:logrelopen}. When extend parallel reduction to
include $\eta$ reduction for functions. This effectively adds $\eta$
laws for functions to our equational theory since coherence, the
untyped relation used for type conversion, is built on top of parallel
reduction.

We can recover the same confluence result about parallel reduction
using the standard techniques from \citet{barendregt, takahashi},
though specifically for the de Bruijn representation, we need the
following anti-renaming lemma about parallel reduction:
\begin{lemma}[Par anti-renaming]
  \label{lemma:parantirenaming}
  If $[[a {xi} => b0]]$, then there exists some $[[b]]$ such that
  $[[b {xi} = b0]]$ and $[[a => b]]$.
\end{lemma}

% The idea of a term blocked from $\beta$ reduction is
% captured by $[[ne]]$, the set of neutral terms. We augment the
% interpretation of booleans and the empty types with terms that
% evaluate to neutral terms. Furthermore, types that are neutral terms
% can also be assigned a meaning since they may be inhabited by other
% neutral terms.
The syntactic forms $[[ne]]$ and $[[nf]]$ capture the neutral terms
and normal forms with respect to $\beta$-reduction, but can still take
$\eta$ reduction steps. Furthermore, we can show that parallel
preserves $[[ne]]$ and $[[nf]]$.

The predicates $[[wne a]]$ and $[[wn
a]]$ describe terms that can evaluate into $\beta$-neutral or
$\beta$-normal form through parallel reduction.

In the definition of the logical relation, we omit the rules for
the function and universe cases since they remain identical to the
original version in Figure~\ref{fig:logrel}.
The changes to \rref{I-Bool} and \rref{I-Void} follow the exact same pattern.
An open term of type $[[Bool]]$ does
not necessarily reduce to $[[true]]$ or $[[false]]$, but may reduce to
a variable, or more generally, a neutral term. Likewise, the
$[[Void]]$ type, while remains uninhabited under an empty context, may
be inherited by the set of neutral terms when there is a variable in
the context that allows us to inhabit $[[Void]]$.

The rule for equality type $[[a ~ b : A]]$ is augmented with the preconditions that
$[[a]]$, $[[b]]$, and $[[A]]$ are all in their normal forms since
otherwise our model would include equality types that are themselves
not normalizing. Furthermore, the side condition $[[a <=> b]]$ is only
available when the equality proof reduces to $[[refl]]$. If the proof
term reduces to a neutral term, then there is nothing we can learn
about the relationship between $[[a]]$ and $[[b]]$.

Finally, in a
non-empty context, a type itself may evaluate to a neutral term and in
turn can only inhabited by neutral terms, thus the addition of \rref{I-Ne}.

Interestingly, all the properties we have shown in
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} up to the
fundamental lemma can be proven in the exact same order, where the new
cases related due to \rref{I-Ne} and the augmentation of neutral terms
to \rref{I-Void, I-Eq, I-Bool} can be immediately discharged by
Lemma~\ref{lemma:parnenf}.

Before we can prove the fundamental theorem and derive the
normalization property as its corollary, we need to additionally
formulate and prove the adequacy property.

Let us first start with some auxiliary lemmas and definitions.
\begin{lemma}[Ext Wn]
  \label{lemma:extwn}
  If $[[wn a i]]$, then $[[wn a]]$.
\end{lemma}
\begin{proof}
  By induction over the length of the reduction sequence in $[[wn a
  i]]$. The proof relies on Lemma~\ref{lemma:parantirenaming} and
  \ref{lemma:parnenf}.
\end{proof}

\begin{lemma}[wne wn]
  \label{lemma:wnewn}
  If $[[wne a]]$ and $[[wn b]]$, then $[[wne a b]]$.
\end{lemma}
\begin{proof}
  Immediate by lexicographical induction over the length of the reduction sequences in
  $[[wne a]]$ and $[[wn b]]$.
\end{proof}

\begin{definition}[CR]
  Let $[[S]]$ be a set of lambda terms. We say that
  \begin{itemize}
  \item $[[S]] \in CR_1$ if $[[wne a implies a in S]]$
  \item $[[S]] \in CR_2$ if $[[a in S implies wn a]]$
  \item $[[S]] \in CR$ if $[[S]] \in CR_1$ and $[[S]] \in CR_2$
  \end{itemize}
\end{definition}

\begin{lemma}[$CR_1$ for type]
  \label{lemma:cr1ty}
  If $[[wne A]]$, then we have $[[Interp I i A {a | wne a }]]$.
\end{lemma}
\begin{proof}
  Immediately from \rref{I-Ne} and \rref{I-Par}.
\end{proof}

\begin{lemma}[$CR$ for interpreted sets]
  \label{lemma:crel}
  If $[[I]] \in CR$ and $[[Interp I i A S]]$, then $[[S]] \in CR$.
\end{lemma}
\begin{proof}
  By induction over the derivation of $[[Interp I i A S]]$. The base
  cases are all immediate.

  In the function case ($[[Pi A B]]$), we use the induction
  hypothesis to show that variables, which are special cases of
  neutral terms, semantically inhabit the input type $[[A]]$. We apply the
  function to an arbitrary variable and we can then use
  Lemma~\ref{lemma:extwn} to conclude the $CR_2$ property.

  To show
  the $CR_1$ property, we need to prove that given a neutral term $[[b]]$
  such that $[[wne b]]$,
  and a term $[[a]]$ that semantically inhabits $[[A]]$, $[[b a]]$ gives
  us a term that semantically inhabits $[[B {a}]]$. From the induction
  hypothesis, $[[A]]$ also satisfies $CR_2$ and therefore $[[wn a]]$
  holds. By Lemma~\ref{lemma:wnewn}, we have $[[wne b a]]$. By $CR_2$
  from the induction hypothesis, since every term that evaluates to
  some neutral form semantically inhabits $[[B {a}]]$. Done.
\end{proof}

\begin{lemma}[$CR_2$ for type]
  \label{lemma:cr2ty}
  If $[[I]] \in CR$ and $[[Interp I i A S]]$, then $[[wn A]]$ holds.
\end{lemma}
\begin{proof}
  By induction over the derivation of $[[Interp I i A S]]$. The
  function case uses Lemma~\ref{lemma:crel} and Lemma~\ref{lemma:extwn}.
\end{proof}

\begin{lemma}[$CR_2$ for types (Rec)]
  \label{lemma:cr2tyrec}
  If $[[InterpR i A S]]$, then $[[wn A]]$.
\end{lemma}
\begin{proof}
  By strong induction over $[[i]]$ and Lemma~\ref{lemma:cr2ty}.
\end{proof}

\begin{lemma}[$CR$ for terms (Rec)]
  \label{lemma:crelrec}
  If $[[InterpR i A S]]$, then $[[S]] \in CR$.
\end{lemma}
\begin{proof}
  After unfolding the definition of $[[InterpR i A S]]$, trivial by
  Lemma~\ref{lemma:cr2ty}, \ref{lemma:crel} and \ref{lemma:cr1ty}.
\end{proof}

The formulation of the valuation and semantic well-typedness from
Figure~\ref{fig:semtyping} and the fundamental lemma remains
unchanged.

The proof of the fundamental lemma is still carried out by induction
over the typing derivation, where the additional neutral term related
cases are handled by Lemmas\ref{lemma:cr2tyrec} and
\ref{lemma:crelrec}.

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{theorem}[Existence of $\beta$-normal form]
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{theorem}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the valuation $[[rho]]$ as the term constructor that injects
  natural numbers as variables. This is
  allowed since Lemma~\ref{lemma:crelrec} says neutral terms
  semantically inhabit any type.
\end{proof}

@@
